### 2020-04-14 manuscript/manuscript.rmd

## Abstract

Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implciations for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from a large number of previous phylogenomic analyses of protein coding genes. These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used. As other authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes. Most other conditions supported Ctenophora-sister. It has been argued that the results using site heterogeneous models with reduced outgroup sampling (which favor Porifera-Sister) should be favored because they best address different frequencies of amino acids found across sites and in more distantly related outgroups. We perform new analyses with these data to characterize the differences between matrices under consistent conditions. Our reanalyses are generally consistent with previous results. We performed additional analyses using new site heterogeneous models in iqtree, and find that, unlike the phylobayes analyes, these support Ctenophora-sister rather than Porifera-sister. This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used. We also quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models). We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic. Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

## Introduction

Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life. However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig. 1) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the placement of these lineages is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns. However, a wide consensus of animal origin has not been reached despite multiple efforts.

@@

![](figures/Figure_overview.png)
**Fig. 1.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta (designated by the blue node). (D) A ctenophore. (E) A sponge.

@@

## Results and Discussion

@@

**Fig. 2.** Matrix taxa and gene sampling. Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

@@

```{r figure_support_analyses_published}
analyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
x1 = factor(analyses_published$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
analyses_published %>%
	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
   xlab("Model complexity") + ylab("Clade")
```
**Fig. 3.**  A total of `r nrow(analyses_published)` analyses were transcribed from the literature (Table XX).

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. 3 and Supplementary Table 2). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models in several matrices, including Philippe2009, Ryan2013_est and Whelan2015. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig. 3). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in most other conditions.

@@

```{r figure_support_analyses_new}
analyses_new=analyses_new %>%
mutate(model=recode(model,
                    `GTR20`="GTR",
                    `CAT+F81`="Poisson+CAT"))

x1 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
analyses_new %>%
	ggplot( aes(x=x1, y=clade, col=result, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
       xlab("Model complexity") + ylab("Clade")
```
**Fig. 4.**  A total of `r nrow(analyses_new)` new phylgoenomic analyses were conducted in this study (Supplementary Table XX).

One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Thus, it easily causes confusions with the increase number of studies.  Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig. 4; Supplementary Table 3). Importantly, we also include a CAT-like, site-heterogeneous C10 -C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.

Here we reanalyzed 17 matrices from 12 studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1). It should be noted that we didn't include GTR+C60 models in model testing since it is computational too expensive with large number of parameters that need to be estimated.  We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig. 4; Supplementary Table 3).

@@

We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig. Extended Data Fig. 3). Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of animal origin is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C60 models in Phillippe2009 and Whelan2017_full matrices.

# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
#                    widths = unit(c(4, 4), "cm"),
#                    heights = unit(c(2, 5), c("in", "lines")))
#
# grid.newpage()
# grid.draw(fig5)
```

**Fig.5** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice. (A). Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Philippe2009 matrix. (B). Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Ryan2013_est matrix. (C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix. The (\Delta$GLS\) values for the genes across each data matrix are provided in Supplementary Table 4.

@@

### Bias of data-recoding
There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies. They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices. Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig. 4). Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi, we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 5). Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).

**Extended Data Fig. 4.** (A) The two alternative hypotheses for deep animal relationships considered here. Relationships that are not part of these hypotheses are shown as unresolved polytomies. (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.

![](figures/Figure_recoding_summary.png)
**Extended Data Fig. 6.** A graphical representation of the posterior probabilities for the 32 analyses presented by Feuda *et al.* in their Table 3. Cells are color coded by whether posterior probability is > 95 for Porifera-sister, > 95 for Ctenophora-sister, or neither (unresolved). Posterior predictive (PP) statistics were estimated for the 16 analyses in the top two rows of this figure (the Chang and full Whelan matrices), but not the bottom two (the Whelan matrices with reduced outgroup sampling).

**Extended Data Fig. 5.** The subset of eight CAT-GTR+G analyses with posterior predictive (PP) scores that is the focus of Feuda *et al.*'s primary conclusions. These are a subset of the 32 analyses presented in their Table 3 and graphically here in the upper right pane. The eight analyses are for two datasets (Chang and Whelan) and four coding schemes. The coding schemes are the original 20 state amino acid data, and three different six state recodings that group amino acids based on different criteria: KGB6, D6, and SR6. Cells are color coded as in Supplemental Fig XXRecoidng_summary. Only one of these analyses, the SR6 coding of the Whelan matrix, has > 95 support for Porifera-sister. The 20-state and 6-state points on the plots in Fig XXRecoidng_summary of the main text correspond to the 20 and SR6 Whelan cells shown here. The presented statistics for these cells are posterior probability of Porifera-sister, Maxdiff (with lower scores indicating better convergence of runs), and five posterior predictive statistics (where lower absolute value indicates better model adequacy). The only one of these eight analyses that provides strong support for Porifera-sister is not the most adequate analysis by any of the posterior predictive scores, and showed the poorest convergence according to Maxdiff.

@@

### Data selection and wrangling

@@

### Matrix comparison and annotation

#### Taxon name reconciliation

@@

#### Sequence Comparisons

@@

#### Partition Network

@@

The 20 state matrices are the same across all analyses since they are not recoded. Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig. 4) give insight into the technical variance of the inference process.

@@

## Supplementary Methods

### Matrix mapping

@@

### Taxa comparison across matrices

@@

>>>>

## Abstract
XX(Current Version)
Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships in the animal tree of life. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implications for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from all previous phylogenomic analyses and perform new analyses with these data to characterize the differences between matrices under consistent conditions. Our reanalyses are generally consistent with previous results. We next quantify the support of phylogenetic signal and conduct sensitive analyses in several representative data matrices. We found that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions. We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address variation in amino acid frequencies, is problematic. Thus, our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

XX(Previous version)
Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships among major animal lineages. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implciations for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from a large number of previous phylogenomic analyses of protein coding genes. These published analyses varied in many regards, including the genes sampled, the outgroup taxa included, the ingroup taxa included, the models of molecular evolution applied to the data, and the phylogenetic methods and software used. As author authors have noted, these studies tend to support Porifera-sister when the outgroups are reduced to only only Choanoflagellata (the closest outgroup to animals), and site heterogeneous CAT models are used in the software tool PhyloBayes. Most other conditions supported Ctenophora-sister. It has been argued that the results using site heterogeneous models with reduced outgroup sampling (which favor Porifera-Sister) should be favored because they best address different frequencies of amino acids found across sites and in more distantly related outgroups. We perform new analyses with these data to characterize the differences between matrices under consistent conditions. Our reanalyses are generally consistent with previous results. We performed additional analyses using new site heterogeneous models in iqtree, and find that, unlike the phylobayes analyes, these support Ctenophora-sister rather than Porifera-sister. This suggests that some of the variation seen previous studies was due to different phylogenetic analysis software, rather than the different models that were used.
We next quantify the support of phylogenetic signal over the two hypotheses and find substantial differences in the distribution of phylogenetic signal between models and outgroup choice (especially CAT models). We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address vaariation in amaino acid frequencies, is problematic. Our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

## Main
Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life. However, over the past decade, there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig. 1) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the placement of these lineages is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for the evolution of complex structures such as cell types, nervous systems, and developmental patterns. However, a wide consensus of animal origin has not been reached despite multiple efforts.

@@

![](figures/Figure_overview.png)
**Fig. XXFigure_overview.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta (designated by the blue node). (D) A ctenophore. (E) A sponge.

@@

@@

**Fig. XXtaxon_rectangles.** Matrix taxa and gene sampling. Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

@@

```{r figure_support_analyses_published}
aanalyses_published=read_tsv("../data_processed/tables/previously_published_analyses.tsv")
analyses_published_filter = dplyr::filter(analyses_published, !grepl('Recoding', model_combined))
x1 = factor(analyses_published_filter$model_combined, levels=c("WAG", "LG", "GTR", "data partitioning", "Recoding + GTR", "Recoding + GTR + CAT", "Poisson + CAT", "GTR + CAT"))
p1 = analyses_published_filter %>%
	ggplot( aes(x=x1, y=clade, col=final, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
   xlab("Model complexity") + ylab("Clade")

analyses_new=analyses_new %>%
mutate(model=recode(model,
                    `GTR20`="GTR",
                    `CAT+F81`="Poisson+CAT"))

x2 = factor(analyses_new$model, levels=c("WAG", "GTR", "Poisson+C60", "WAG+C60", "LG+C60", "Poisson+CAT"))
p2 = analyses_new %>%
	ggplot( aes(x=x2, y=clade, col=result, shape=inference) ) +
		geom_jitter( width = 0.20, height = 0.20, alpha=0.7, size=2 ) +
		scale_colour_manual(values = result_colors) +
		theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))+
       xlab("Model complexity") + ylab("Clade")

ggarrange(p1, p2, ncol = 2, labels = c("A", "B"), common.legend = TRUE, legend = "right")

```
**Fig. XXfigure_support_analyses.**  A summary of phylogenomic results from the previous studies and reanalyses conducted in this study. (A) A total of `r nrow(analyses_published_filter)` analyses were transcribed from the literature (Table XX). (B) A total of `r nrow(analyses_new)` new phylogenomic analyses were conducted in this study (Supplementary Table XX).

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. 3A and Supplementary Table 2). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models in several matrices, including Philippe2009, Ryan2013_est and Whelan2015. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig. 3A). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in most other conditions.

@@

One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Thus, it easily causes confusions with the increase number of studies.  Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig. 3B; Supplementary Table 3). Importantly, we also include a CAT-like, site-heterogeneous C10 -C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.

Here we reanalyzed 17 matrices from 12 studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1). It should be noted that we didn't include GTR+C60 models in model testing since it is computational too expensive with large number of parameters that need to be estimated.  We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig. 3B; Supplementary Table 3).

@@

We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig. Extended Data Fig. 3A). Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of animal origin is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C60 models in Phillippe2009 and Whelan2017_full matrices.

@@

``` {r sensitive, fig.width=8, fig.height=8}
sensitive = read_tsv("/animal_root/data_processed/tables/sensitive.tsv")
Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_choanozoa" )
Philippe2009$Model = factor(Philippe2009$Model, levels=c("Poisson_CAT60" , "Poisson_CAT90","Poisson_CAT120","Poisson_CAT150","Poisson_CAT180"))
p1<-ggplot(Philippe2009, aes(Model))+
  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Choanozoa")

Philippe2009 = sensitive %>% filter(Matrix == "Philippe2009_only_Holozoa")
Philippe2009$Model = factor(Philippe2009$Model, levels=c("Poisson_CAT360", "Poisson_CAT400","Poisson_CAT440","Poisson_CAT480"))
p2<-ggplot(Philippe2009, aes(Model))+
  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=2), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Philippe2009_Holozoa")

Whelan2017_full_only_choanozoa = sensitive %>% filter(Matrix == "Whelan2017_full_only_choanozoa")
Whelan2017_full_only_choanozoa$Model = factor(Whelan2017_full_only_choanozoa$Model, levels=c("Poisson_CAT60" , "Poisson_CAT340","Poisson_CAT380","Poisson_CAT420","Poisson_CAT460"))
p3<-ggplot(Whelan2017_full_only_choanozoa, aes(Model))+
  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(x ="Number of CAT substitutional categories", y = "Support") + ggtitle("Whelan2017_full_Choanozoa")

Whelan2017_strict = sensitive %>% filter(Matrix == "Whelan2017_strict" )
Whelan2017_strict$Model = factor(Whelan2017_strict$Model, levels=c("Poisson_CAT60" , "Poisson_CAT70","Poisson_CAT80","Poisson_CAT90","Poisson_CAT100","Poisson_CAT110", "Poisson_CAT120","Poisson_CAT150","Poisson_CAT180", "Poisson_CAT360"))
p4<-ggplot(Whelan2017_strict, aes(Model))+
  geom_line(aes(y = `Ctenophora-sister`, colour = "Ctenophora-sister", group=1), color=("#E69F00")) + geom_point(aes(y = `Ctenophora-sister` ), color=("#E69F00"))+
  geom_line(aes(y = `Porifera-sister`, colour = "Porifera-sister",group=1), color=("#009E73")) + geom_point(aes(y = `Porifera-sister` ), color=("#009E73"))+
  scale_fill_manual(values=c("#E69F00", "#ebd196")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(x ="Number of CAT substitutional categories", y = "Support")+ ggtitle("Whelan2017_strict")

ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")),nrow = 3,
          p3,p4, labels = c("C", "D"))
```
**Fig. XXsensitive.**

### Bias of data-recoding
There is growing interest in data-recoding and previous study [@feuda2017improved] hoped that data-recoding would reduce potential artifacts due to differences across species in amino acid frequencies. They report that posterior predictive (PP) analyses [@Bollback:2002to] indicate 6-state recoded analyses have better model adequacy than 20-state amino acid analyses, and "Porifera-sister was favored under all recoding strategies" in Whelan2015_D20 and Chang2015 data matrices. Here we focus on two aspects of Feuda *et al.* First, we found that many of their recoded analyses are actually unresolved (*i.e.*, without strong support for either Porifera-sister or Ctenophora-sister), and that the analyses with the best posterior predictive scores do not provide strong support for Porifera-sister (Extended Data Fig. 3B). Second, we created four new random recoding schemes by shuffling the amino acids in the SR-6 scheme (see methods). When we applied each of these randomized codes to the Whelan matrix and analyzed them under the CAT-GTR+G model with phylobayes-mpi, we observed similar results as for the empirical SR-6 recoding. Like SR-6 recoding, random recoding increases support for Porifera-sister and improves the apparent adequacy of models to explain heterogeneity of states across taxa (PP taxon hetero mean and max, Extended Data Fig. 5). Thus, our analyses show the impact of recoding is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation (see more detailed discussion in supplementary text).

@@

### Data selection and wrangling (Ben)

@@

### Matrix comparison and annotation (Ben)

#### Taxon name reconciliation(Ben)

@@

#### Sequence Comparisons (Ben)

@@

#### Partition Network (Ben)

@@

The 20 state matrices are the same across all analyses since they are not recoded. Since all the 20 state matrices are the same, variation between 20-state results (as in the left side of each pane of Extended Data Fig. 3B) give insight into the technical variance of the inference process.

@@

### Matrix mapping (Ben)

@@

### Taxa comparison across matrices (Ben)

@@

**Extended Data Fig. XXfigure_gene_composition.** Annotation and representation of BUSCO and ribosomal genes in each data matrix. (A). The number of partitions with BUSCO annotations in each matrix, relative to the number of partitions. (B). The percentage BUSCO annotations in each matrix. (C). The number of partitions with annotations of ribosomal genes in each matrix, relative to the number of partitions. (D). The percentage  ribosomal genes in each matrix.

**Extended Data Fig. XXfigure_matrix_overlap.** Pairwise overlap between each of the primary matrices considered here. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled. The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.

![](figures/Figure_Pairwise_heatmap_animal_root.png)
**Extended Data Fig. XXpairwise_heatmap.** Pairwise intersections of the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices calculated by Euclidean distance. (A) Pairwise comparisons across genes that supported Ctenophora-sister in different analyses. (B) Pairwise comparisons across genes that supported Porifera-sister in different analyses. The color scale indicates the Pearson correlation, where 1 is positive linear correlation, 0 is no linear correlation, and −1 is negative linear correlation. The pairwise intersection heatmap is generated by Intervene [@khan2017intervene].

# fig5 <- gtable_matrix(name = "demo", grobs = grob_grid,
#                    widths = unit(c(4, 4), "cm"),
#                    heights = unit(c(2, 5), c("in", "lines")))
#
# grid.newpage()
# grid.draw(fig5)
```
**Extended Data Fig. XXphylogetic signal** The distribution of phylogenetic signal for two alternative topological hypotheses on the earliest-branching animal lineage with different models and outgroup choice. (A). Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Philippe2009 matrix. (B). Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Ryan2013_est matrix. (C) Proportion of genes supporting each of two alternative hypotheses for each of two alternative hypotheses for Whelan2017_full matrix. The (\Delta$GLS\) values for the genes across each data matrix are provided in Supplementary Table 4.

**Extended Data Fig. XXrandom_recoding.** (A) The two alternative hypotheses for deep animal relationships considered here. Relationships that are not part of these hypotheses are shown as unresolved polytomies. (B) Each of the six plots presents one statistic, which include Posterior probability of Porifera-sister and the five Posterior Predictive (PP) statistics considered by Feuda *et al.* Within each plot, there are six lines for six different analyses. These six analyses are the published SR-6 analyses presented by Feuda *et al.* (SR6-published), analyses obtained by applying the same methods to the same data to to confirm that I can reproduce their published results (SR6-reproduced), and four analyses based on randomized recoding matrices obtained by shuffling the SR-6 coding scheme (random-00 – random-03). Each analysis includes results for 20 states (the raw amino acid data, shown by the left point) and for 6 states (the 6-state recoded data, shown by the right point). For each statistic, the results obtained with the random recoding are similar to those of the SR6 recoding. This indicates that the impact of recoding is dominated by discarding data when collapsing from 20 states to 6 states, not accommodating compositional heterogeneity across lineages.
